Taking Dyads Seriously

### 2021-02-03 paper/2_model.tex
push up memo revisions

When modeling dyadic data scholars typically employ a generalized linear model (GLM) estimated via maximum-likelihood. This type of model is expressed via a stochastic and systematic component \citep{pawitan:2013}. The stochastic component reflects our assumptions about the probability distribution from which the data are generated: $y_{ij} \sim P(Y | \theta_{ij})$, with a probability density or mass function, such as the normal, binomial, or Poisson, and we assume that each dyad in the sample is independently drawn from that particular distribution, given $\theta_{ij}$. The systematic component characterizes the model for the parameters of that distribution and describes how $\theta_{ij}$ varies as a function of a set of nodal and dyadic covariatse, $\mathbf{X_{ij}}$: $\theta_{ij} = \bm\beta^{T} \mathbf{X}_{ij}$. A fundamental assumption we make when applying this modeling technique is that given $\mathbf{X}_{ij}$ and the parameters of our distribution, each of the dyadic observations are conditionally independent. The importance of this assumption becomes clearer in the process of estimating a GLM via maximum likelihood. After having chosen a set of covariates and specifying a distribution, we construct the joint density function over all dyads, for example:

@@

The LFM contribution to the AME is in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. The construction of the LFM here is actually quite similar to work on low rank approximations in computer science and has been applied to the development of recommender systems that companies like Amazon and Netflix use to model customer behavior \citep{resnick:varian:1997,bennett:lanning:2007}.\footnote{The LFM also shares similarities with work in the econometric literature on interactive fixed effects \citep{bai:2009,pang:2014}. In this stream of work, interactive fixed effects are used to deal with cross-sectional dependence in TSCS data, in such a way that a latent factor for time can be used to capture common shocks to actors and a latent factor on actors can capture varying responses to those shocks.} This model posits a latent vector of characteristics
$\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When computing the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.} Note that this model easily generalizes to the case, common in IR, where interactions are undirected (for example the presence of conflict or a bilateral investment treaty). In the case of the SRRM, $\rho$ is constrained to be one and instead of separate sender and receiver random effects a single actor random effect is utilized. For the LFM, an eigen-decomposition scheme is used to capture higher-order dependence patterns. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data. \footnote{Parameter estimation in the AME takes place within the context of a Gibbs sampler in which we iteratively sample from the posterior distribution of the full conditionals for each parameter. The algorithm is provided in the Appendix.}

>>>>

When modeling dyadic data scholars typically employ a GLM estimated via maximum-likelihood. This type of model is expressed via a stochastic and systematic component \citep{pawitan:2013}. The stochastic component reflects our assumptions about the probability distribution from which the data are generated: $y_{ij} \sim P(Y | \theta_{ij})$, with a probability density or mass function, such as the normal, binomial, or Poisson, and we assume that each dyad in the sample is independently drawn from that particular distribution, given $\theta_{ij}$. The systematic component characterizes the model for the parameters of that distribution and describes how $\theta_{ij}$ varies as a function of a set of nodal and dyadic covariatse, $\mathbf{X_{ij}}$: $\theta_{ij} = \bm\beta^{T} \mathbf{X}_{ij}$. A fundamental assumption we make when applying this modeling technique is that given $\mathbf{X}_{ij}$ and the parameters of our distribution, each of the dyadic observations are conditionally independent. The importance of this assumption becomes clearer in the process of estimating a GLM via maximum likelihood. After having chosen a set of covariates and specifying a distribution, we construct the joint density function over all dyads, for example:

@@

The LFM contribution to the AME is in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. The construction of the LFM here is actually quite similar to work on low rank approximations in computer science and has been applied to the development of recommender systems that companies like Amazon and Netflix use to model customer behavior \citep{resnick:varian:1997,bennett:lanning:2007}.\footnote{The LFM also shares similarities with work in the econometric literature on interactive fixed effects \citep{bai:2009,pang:2014}. In this stream of work, interactive fixed effects are used to deal with cross-sectional dependence in TSCS data, in such a way that a latent factor for time can be used to capture common shocks to actors and a latent factor on actors can capture varying responses to those shocks.} This model posits a latent vector of characteristics
$\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When computing the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.} Note that this model easily generalizes to the case, common in IR, where interactions are undirected (for example the presence of conflict or a bilateral investment treaty). In the case of the SRRM, $\rho$ is constrained to be one and instead of separate sender and receiver random effects a single actor random effect is utilized. For the LFM, an eigen-decomposition scheme is used to capture higher-order dependence patterns. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data. Parameter estimation in the AME takes place within the context of a Gibbs sampler in which we iteratively sample from the posterior distribution of the full conditionals for each parameter.\footnote{Additional details on the sampling procedure and the full conditional distributions can be found in: \citet{hoff:2005,hoff:2008,minhas:etal:2019,hoff:2021}.}
